---
title: 'Length-based Processes in an Age-Structured Model'
author: ''
date: '2022-05-02'
slug: len-age-conv
categories: []
tags: []
subtitle: ''
summary: ''
authors: []
lastmod: '2022-05-02T13:58:32-07:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

This is in the category of "stuff I thought I understood" until I didn't.

Let's say you have an age-based model with integer-year bins `a` thru `A+`. The model is going to keep track of the dynamics of the fish population as they step through each age, but (annoyingly, perhaps) some of your modeled processes are *length*-based. This can occur when you are working with lots of different fleets, for which gear selectivity is better specified by length, or even the simple inclusion of our favorite $w_{l} = a l^b$ calculation, which will require a notion of a fish length-at-age to play nicely within an age-structured model.

Converting these length-based estimates (or quantities) into age-based estimates introduces a problem because of **uncertainty in length-at-age**, in other words, the fact that fish of length $l$ could be of various ages $a$. Let's make up some data:

```{r, setup, include=FALSE}
require(ggplot2)
require(dplyr)
theme_set(ggsidekick::theme_sleek())
options(dplyr.summarise.inform = FALSE)
```

```{r echo = T, warning = FALSE, message = FALSE}
## check out my post "Why do we use ln(19)?" for more info on this syntax:
logistic <- function(bin, b50, b75){
  selage <- 1/(1+exp(-(log(15))*(bin-b50)/(b75-b50)))
  return(selage)
} 

age_bins <- 0:50
length_bins <- 1:65
sigma_length <- 7.16

## a logistic, length-based selectivity curve
length_selex <- sapply(length_bins, logistic, b50 = 30, b75 = 52)

```
Here's our selectivity curve; there's a vertical line at `l = 28cm` which we'll use for reference throughout this post.
```{r echo = FALSE, warning = FALSE, message = FALSE}
ggplot(data.frame(cbind(length_selex,length_bins)),
       aes(x = length_bins, y = length_selex)) +
  geom_line(lwd = 1.1, col = 'blue') +
  geom_vline(xintercept = 28, linetype = 'dotted')+
  labs(x = 'length', y= 'proportion selected', main = 'A length-based logistic selectivity curve with L_50% = 30cm and L_75% = 52cm')
```
So let's say we're interested in the selectivity at `l = 28cm`. In a length-based population dynamics model, we'd be set to just return the selectivity at `l = 28cm` for any fish within that respective bin (in 1-cm bins, fish lengths greater than 27cm and less than or equal to 28cm would all be assigned this value.) 

However, in an age-based model, we are instead interested in translating this curve into the expected selectivity-at-*age*. And given age $a$, a fish could be a broad range of lengths $l$, making it less clear which value we would "read off" the blue curve above. This variation is caused by the variable $\sigma$, which I've set at ~7cm -- this reflects the standard deviation in length-at-age for our stock. 

Here's how it looks in practice. First, we can generate our vonB-based expected length-at-age (here I assume $t_0$ is 0, and am plotting the observed variation for 500 datasets, which should approximate the value for `sigma_length` specified above). 


```{r message = FALSE, warning = FALSE}
vonb <- function(age, linf, kappa, sigma_length){
  len <- linf*(1-exp(-kappa*(age)))+rnorm(1,0,sigma_length)
}

## simulate some datasets
length_at_age = matrix(NA, nrow = length(age_bins), ncol = 500)
for(i in 1:500){
  length_at_age[,i] <- sapply(age_bins, vonb, linf = 44,
                              kappa = 0.15, sigma_length)
}
length_at_age <- bind_cols(age = age_bins, 
                           meanL = rowMeans(data.frame(length_at_age)),
                           sdL = apply(data.frame(length_at_age),1,sd))%>%
  mutate(lci = meanL-1.96*sdL, uci = meanL+1.96*sdL)

ggplot(length_at_age,
       aes(x = age_bins)) +
  geom_line(lwd = 1.1, col = 'grey', aes(y = meanL)) +
  geom_ribbon(aes(ymin = lci, ymax = uci), fill = alpha('grey',0.3))+
  geom_hline(yintercept = 28, linetype = 'dotted')+
  labs(x = 'age', y= 'length')
```

**Notice how the horizontal line at 28cm corresponds to multiple ages within the confidence interval?** How do we account for the fact that there is a chance that the the selectivity applicable to a given age can be characterized by the selectivity at several lengths, and how do we weight the applicability of that selectivity value given what we know about growth? Enter $\psi_{l,a}$, the probability of being in length bin $l$ at age $a$, a.k.a. the normal distribution $\phi$ of size-at-age. 

$$
\begin{align}
\psi_{al} &= \begin{array} 
   & \Phi (1,\tilde l_{a}, \sigma)) & if \space l = 0cm \\
   \Phi (l+1,\tilde l_{a}, \sigma)) - \Phi (l,\tilde l, \sigma))& if \space 0cm, < l < L+ \\
   1-\Phi (l,\tilde l_{a}, \sigma)) & if \space l = L+\\
\end{array}
\end{align}
$$

Where $l$ is the length bin in question, $\tilde l$ is the observed length-at-age, and  $\sigma$ is the variation in growth. If we simply used a "lookup" method (which I have certainly done in a pinch!), and assigned the selectivity at age $a$ to the selectivity at the expected length at age $a$, we'd be introducing bias. 


```{r message = FALSE, warning = FALSE}
## this function returns the distribution of expected length-at-age 
## given a bin and observed length-at-age
getP_al <- function(len_bin, len_obs, length_sigma){
  pal = NA
  if(len_bin == 1){
    pal <- pnorm(len_bin, len_obs, length_sigma)
  } else if (len_bin > min(length_bins) & len_bin < max(length_bins)){
        pal <- pnorm(len_bin+1, len_obs, length_sigma) - 
          pnorm(len_bin, len_obs, length_sigma)
  } else{
        pal <- 1-pnorm(len_bin, len_obs, length_sigma)
  }
  pal
}
PAL <- matrix(NA, nrow = length(age_bins), ncol =  length(length_bins))
for(a in age_bins){
  for(l in length_bins){
    ## instead of using means, could re-do vonB with error turned off
    PAL[a,l] <- getP_al(len_bin=l,len_obs=length_at_age$meanL[a],sigma_length )
  }
}
longPal <- data.frame(PAL) %>%
  mutate(age_bins) %>%
  reshape2::melt(id = 'age_bins') %>% 
  mutate(len =as.numeric(gsub("X","",variable))) 
```
Here's a subset of what $\psi_{al}$ looks like for ages 5-10. Notice (again) how length = 28cm crosses multiple age curves. The good news is, $\psi_{al}$ has given us a sense of the relative probability of being age $a$ given length $l$. (It looks like $l = 28cm$ is most likely a fish of age 7 or 8).

```{r echo = F}
ggplot() +
  labs(y = 'P(length|age)',  
       color = 'Age', x = 'Length')+
  theme(legend.position = 'right')+
  scale_color_brewer(palette = 'Spectral')+
  geom_line(data = subset(longPal, age_bins %in% 5:10),lwd = 1,
            aes(x = len, y = value, color = factor(age_bins) )) +
  geom_vline(xintercept = 28, lty = 'dotted')
```
To put this all together, we need to compute the dot-product of the selectivity-at-length and the expected distribution of length-at-age.



In math, we're doing this:
$$
Sel_a = \sum_l Sel_l \cdot \psi_{al}
$$




Which is really just a cheap way of approximating the correct estimator of the expected selectivity of an animal at age $a$: 

$$
Sel_a = \int_L Sel_L  \psi_{aL} dL
$$

In words: for each length bin $l$, multiply the selectivity at that length times the probability of being at age $a$ given length $l$. Sum these quantities across all length bins for each age bin.

In pictures: let's say we are interested in $a = 5$. We are going to multiply each point on the length-selectivity by its corresponding point (by color and ID number) on the probability of length-at-age-5 plot, then sum the results. 

```{r echo = FALSE}
ggplot(data.frame(cbind(length_selex,length_bins))%>% mutate(ID = 1:nrow(.)),
       aes(x = length_bins, y = length_selex)) +
  theme(legend.position = 'none')+
  scale_color_grey() +
  geom_text(aes(x = length_bins, y = length_selex,  label = ID,col = factor(length_bins)))+
  labs(x = 'length', y= 'proportion selected' )
```

```{r echo = FALSE}
ggplot() +
  labs(y= 'P(length|age)',
       color = 'Age', x = 'Length')+
  theme(legend.position = 'none')+
  scale_color_grey() +
  geom_text(data = subset(longPal, age_bins ==5) %>% mutate(ID = 1:nrow(.)),
            aes(x = len, y = value, label = ID, col = factor(len) )) +
  geom_vline(xintercept = 28, lty = 'dotted')
```


In code (still using $a = 5$ as an example):

```{r message = FALSE, warning = FALSE} 
sel_age <- merge(longPal,data.frame(cbind(length_selex,length_bins)),
                 by.x = 'len', by.y = 'length_bins') %>%
  ## inner product for each age-length combo
  mutate(prod1 = length_selex*value) %>% 
  group_by(age_bins, len) %>%
  summarise(dotProd = sum(prod1)) %>%
  ungroup() %>%
  group_by(age_bins) %>%
  summarise(dotProdsum = sum(dotProd))

ggplot(sel_age,
       aes(x = age_bins, y = dotProdsum)) +
  geom_line(lwd = 1.1, col = 'grey22') +
  labs(x = 'age', y= 'proportion selected', 
       main = 'Converted from length-at-age')

```

